---
title: 'Reproducible Research: Project Assignment 1'
author: "Alessandro Spilotros"
date: "10 gennaio 2016"
output: html_document
---
Abstract
--------------
This document presents the results from Project Assignment 1 in the Coursera course Reproducible Research. It was generated using R Markdown and generating a HTML file as output.

For the assignment we use data from a personal activity monitoring device which collects data at 5 minute intervals during the day. The data cover two months of activity of an anonymous individual and include the number of steps taken in 5 minute intervals per day.

Loading packages
---------------------------
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```
Loading and tyding the data
---------------------------
```{r}
data <- read.csv("activity.csv", header = TRUE, sep = ',')
```
Using lubridate package to convert the date into a date format in R
```{r}
data$date <- ymd(data$date)
head(data)
```
What is mean total number of steps taken per day?
---------------------------------------------------
For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
3. Calculate and report the mean and median of the total number of steps taken per day

Part 1.
```{r}
StepsPerDay <- data %>%
  filter(!is.na(steps)) %>%
  group_by(date) %>%
  summarize(steps = sum(steps))
head(StepsPerDay)
```
Part 2.
```{r, echo=TRUE}
hist(StepsPerDay$steps, 15, xlim =c(-1000,23000), ylim=c(-1,20), xlab = "Steps per day", ylab = "Frequency", main = "Histogram Steps per Day", col = 'green')
```
Part 3.
```{r}
mean_steps <- mean(StepsPerDay$steps, na.rm = TRUE)
median_steps <- median(StepsPerDay$steps, na.rm = TRUE)
print("mean steps=")
mean_steps
print("median steps=")
median_steps
```

What is the average daily activity pattern?
-------------------------------------------------
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Part 1. 
```{r}
Interval <- data %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(steps = mean(steps))
head(Interval)
```
Part 2.
```{r, echo=TRUE}
plot(Interval$interval,Interval$steps,type = 'l', xlab = 'interval', ylab='steps', col='red')
```
which.max function finds the maximum steps on average across all days
```{r, echo=TRUE}
Interval[which.max(Interval$steps),]
```
The interval found is 835 with 206 steps.

Imputing missing values
-------------------------------------------------------------
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
4. Make a histogram of the total number of steps taken each day and calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Part 1.
finding the total number of missing values
```{r, echo=TRUE}
sum(is.na(data$steps))
```
Part 2-3.
Strategy to fill in the missing values=replacing with the average over the 5 min interval.
```{r, echo=TRUE}
data2 <- data
missing <- is.na(data2$steps)
interval_avg <- tapply(data2$steps, data2$interval, mean, na.rm=TRUE, simplify=TRUE)
data2$steps[missing] <- interval_avg[as.character(data2$interval[missing])]
sum(is.na(data2$steps))
```
Part 4.
Calculating the number of steps taken in each 5-minute interval per day
```{r, echo=TRUE}
steps2 <- data2 %>%
  group_by(date) %>%
  summarize(steps = sum(steps))
head(steps2)
``` 
Plotting the new histogram
```{r, echo=TRUE}
hist(steps2$steps, 15, xlim =c(-1000,23000), ylim=c(-1,25), xlab = "Steps per day", ylab = "Frequency", main = "Histogram Steps per Day (including missing values)", col = 'green')
```
Calculate the mean and median steps with the filled in values:
```{r, echo=TRUE}
mean_steps_2 <- mean(steps2$steps, na.rm = TRUE)
median_steps_2 <- median(steps2$steps, na.rm = TRUE)
mean_steps_2
median_steps_2
```

Including the missing values the median and the mean are now both equal to 10766.19

Are there differences in activity patterns between weekdays and weekends?
---------------------------------------------------------------------------
For this part the weekdays() will come handy. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

